%%%%%
%%
%% Sample document ``thesis.tex''
%%
%% Version: v0.2
%% Authors: Jean Martina, Rok Strnisa, Matej Urbas
%% Date: 30/07/2008
%%
%% Copyright (c) 2008-2011, Rok Strniša, Jean Martina, Matej Urbas
%% License: Simplified BSD License
%% License file: ./License
%% Original License URL: http://www.freebsd.org/copyright/freebsd-license.html
%%%%%

% Available documentclass options:
%
%   <all `report` document class options, e.g.: `a5paper`>
%   withindex   - enables the index. New index entries can be added through `\index{my entry}`
%   glossary    - enables the glossary.
%   techreport  - typesets the thesis in the technical report format.
%   firstyr     - formats the document as a first-year report.
%   times       - uses the `Times` font.
%   backrefs    - add back references in the Bibliography section
%
% For more info see `README.md`
\documentclass[withindex,glossary]{cam-thesis}

% Citations using numbers
\usepackage[numbers]{natbib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Thesis meta-information
%%

%% The title of the thesis:
\title{Executable separation-logic\\
specifications}

%% The full name of the author (e.g.: James Smith):
\author{Rini Banerjee}

%% College affiliation:
\college{Newnham College}

%% College shield [optional]:
% \collegeshield{CollegeShields/Christs}
% \collegeshield{CollegeShields/Churchill}
% \collegeshield{CollegeShields/Clare}
% \collegeshield{CollegeShields/ClareHall}
% \collegeshield{CollegeShields/CorpusChristi}
% \collegeshield{CollegeShields/Darwin}
% \collegeshield{CollegeShields/Downing}
% \collegeshield{CollegeShields/Emmanuel}
% \collegeshield{CollegeShields/Fitzwilliam}
% \collegeshield{CollegeShields/Girton}
% \collegeshield{CollegeShields/GonCaius}
% \collegeshield{CollegeShields/Homerton}
% \collegeshield{CollegeShields/HughesHall}
% \collegeshield{CollegeShields/Jesus}
% \collegeshield{CollegeShields/Kings}
% \collegeshield{CollegeShields/LucyCavendish}
% \collegeshield{CollegeShields/Magdalene}
% \collegeshield{CollegeShields/MurrayEdwards}
\collegeshield{CollegeShields/Newnham}
% \collegeshield{CollegeShields/Pembroke}
% \collegeshield{CollegeShields/Peterhouse}
% \collegeshield{CollegeShields/Queens}
% \collegeshield{CollegeShields/Robinson}
% \collegeshield{CollegeShields/Selwyn}
% \collegeshield{CollegeShields/SidneySussex}
% \collegeshield{CollegeShields/StCatharines}
% \collegeshield{CollegeShields/StEdmunds}
% \collegeshield{CollegeShields/StJohns}
% \collegeshield{CollegeShields/Trinity}
% \collegeshield{CollegeShields/TrinityHall}
% \collegeshield{CollegeShields/Wolfson}
% \collegeshield{CollegeShields/CUniNoText}
% \collegeshield{CollegeShields/FitzwilliamRed}

%% Submission date [optional]:
% \submissiondate{November, 2042}

%% You can redefine the submission notice [optional]:
% \submissionnotice{A badass thesis submitted on time for the Degree of PhD}

%% Declaration date:
\date{My Month, My Year}

%% PDF meta-info:
\subjectline{Computer Science}
\keywords{one two three}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Abstract:
%%
\abstract{%
  My abstract ...
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Acknowledgements:
%%
\acknowledgements{%
  My acknowledgements ...
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Glossary [optional]:
%%
\newglossaryentry{HOL}{
    name=HOL,
    description={Higher-order logic}
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Contents:
%%
\begin{document}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page, abstract, declaration etc.:
%% -    the title page (is automatically omitted in the technical report mode).
\frontmatter{}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Thesis body:
%%
% \chapter{[To delete] Outline}

% \begin{enumerate}
%   \item Introduction
%   \item Background + related work
%   \item Motivating example, including partial + fragmentary specifications
%   \begin{enumerate}
%     \item Meld previous simpler CN spec example with narrative from PLDI submission overview section
%   \end{enumerate}
%   \item Formalisation (Dhruv) -- incl. unpublished partial spec formalisation by Hiro?
%   \begin{enumerate}
%     \item Formalisation of Fulminate's dynamic ownership tracking scheme
%     \item Relationship between that and explicit resource passing scheme
%   \end{enumerate}
%   \item Implementation
%   \begin{enumerate}
%     \item Architecture (incl. diagram from Fulm paper, or something similar). Has diverged a bit now as CN and Fulm wf checks are slightly different.
%     \item Translation scheme + Fulminate runtime library
%     \begin{enumerate}
%       \item CN types $\rightarrow$ C boxed representations. Everything is a pointer (and that's not bad)
%       \item 
%     \end{enumerate}
%   \end{enumerate}
%   % \item Future work
% \end{enumerate}

\chapter{Introduction}

PS 29/04: Intro can be long. Can have the Code-Spec-Test-Debug-Prove diagram! Doesn't need to be short and sweet if that's not how I want it to be.

\section{Thesis statement}

PS 29/04: Might need to come later after motivation.
\section{Motivation}
\section{Testing specifications}

Some short text on what it means to test specifications and how the level of assurance and what it is ensuring differs from their conventional use in verification. (i.e. testing checks properties hold at those particular program points for concrete test cases -- can be used in conjunction with random test generation tooling -- whereas in proof, preconditions imply postconditions etc).

Mention history of testing specifications, dating back to 1970s -- Euclid, Anna.

\subsection{Why test specifications?}

Slide from LFCS/POPL talk -- flesh out. Also see first-year report, ``1.2 Objectives''.

Code-Specify-Test-Debug-Prove? Maybe allude to it here (testing-and-debugging-specs-centred workflow), and have a separate chapter on it later (how Fulminate is used for that cycle and how it's amenable to it).

Comment on how this is not done much at all, and about the social gap between testing and verification communities.


\subsection{other notes}

\begin{enumerate}
  \item in defense of testing (dear formal methods community)
  \item in defense of specification, checking richer mathematical properties, and possibly working up towards verification (dear software engineering and testing communities)
  \item see PS specification paper, useful insights (and credit)
  \item Differing correctness guarantees, and classic cost-benefit comparison
  \item Social gap
  \item Testing aids verification -- gradual assurance slope
  \item Testing is useful even without verification!
  \item NK 29/04: assertions that developers write in C usually only involve C-local variables; they care about performance a lot more than we do, and there's no abstraction/building of ghost state. Therefore, much shallower properties are checked on average. On the other hand, CN describes everything abstractly for the purpose of proof. Fulminate falls somewhere in between, and was not designed to be high-performance from the outset (see: ghost state construction and its overhead), but for running in real systems environments (e.g. EL2) this is something we care about.    
\end{enumerate}

\section{A testing-centred workflow for writing specifications}

Code-Specify-Test-Debug-Prove. Generic here; can make concrete later with CN toolchain for C code.

Include diagram from Keynote slides. Make diagram more generic; our work has been without the Code part of the cycle, so I made diagram specific to the retroactive case, but it needs to be more general. (Diagram currently starts with ``Write partial specs'' which assumes a pre-existing codebase)

\section{Testing separation-logic specifications}

Focus now on systems setting. Motivate why reliability of systems software so important (bug propagating upward, affecting all software running on top of it), and say most systems software is written in C. Look at Kayvan thesis related work briefly?

\subsection{Separation logic}

History:

Hoare logic not good for reasoning about aliasing pointers. Number of inequalities blows up. Motivated creation of separation logic by Reynolds and O'Hearn. $p \mapsto v$, $A \ast B$, frame rule. See LFCS/Birmingham slides. Active research area, lots of research in it today. Link back to systems setting -- these operators make heap memory reasoning compositional which is great when reasoning about programs written in e.g. C or Rust.

\subsection{Executing separation logic}

Comment on technical difficulty of executing $\ast$ operator and give example of how it blows up. Backtracking, arbitrary quantification over pointers, etc.



\section{Thesis outline}
\section{Publications} 
(incl. publications, collaborators -- maybe a few sentences, doesn't need its own subsection)

PS 29/04: Thesis outline and publications might become a single section. Plus discussion of exactly which bits I did.





\chapter{Related work}

See first-year report as a starting point.

\begin{enumerate}
  \item Executable specifications
  \begin{enumerate}
    \item Hoare Logic runtime specification checking (incl. E-ACSL) - no ownership
    \item Linear logic runtime checking - Limin Jia PhD thesis - not SL
    \item Contract checking literature
    \item SLICK (Java runtime SL checking) - not C
  \end{enumerate}
  \item C separation-logic verification tooling (similar tools but specs not executable or not applicable to real-world C code) -- e.g. RefinedC, Gillian. See CN and Fulminate papers' related work sections
  \item Gradual verification -- e.g. Gradual C0
\end{enumerate}



\chapter{Example}

Question for PS and NK: cannot work out whether to present in context of Code-Spec-Test-Debug-Prove from the start, or whether to present complete specifications in the style of proof initially, just to show the mechanism of how Fulminate works, and then only introduce partial specs in later chapter. 

\section{The CN separation-logic specification language}

Cover $p \mapsto v$ and \texttt{Owned}
IntList example (should be in above spec example) and side-by-side comparison between standard SL formulae and CN representation. 


\section{Writing partial CN specifications using Fulminate}

Assuming we have a pre-existing C function to specify in this case.

Small spec example, using IntList. Explain CN semicolon is $\ast$. 

Say Fulminate takes CN+C files, gives C file with checks in the expected places. Fulminate tracks ownership of stack-local and heap addresses; instruments access checks to ensure ownership of an address has been asserted properly before it is read from or written to.

Show Fulminate runtime ownership error from partial spec, debug and fix up spec, run again.


\chapter{Formalisation}

Fulminate paper, link to Dhruv's supplementary material and credit him (+ Hiro if including partial spec formalisation).

\chapter{Implementation}

Fulminate paper plus whatever comes to mind.

\section{Partial specifications}

\begin{enumerate}
  \item Partial function-level specs
  \item Fragmentary specs
  \item Wildcard ownership
\end{enumerate}




\chapter{Case studies}

PS 29/04: Calling this section ``Evaluation'' is kind of underselling the importance of this chapter.

\begin{itemize}
  \item Tutorial examples (CN, Verifast, incremental stack example)
  \item pKVM
  \begin{itemize}
    \item what is pKVM
    \item buddy allocator
    \item hyp allocator
  \end{itemize}
\end{itemize}



% \section{Applying Code-Specify-Test-Debug-Prove}

% \section{Performance}

% copy-paste from PLDI '26 paper

% \section{Bugs}

% Specification bugs; Hiro's experience section from PLDI paper.

% \section{Usability}

% Say upfront we don't have anything sophisticated like user studies. (check status of DARPA ones)

\begin{itemize}
  \item integration with LLDB
  \item error messages (TODO: improve + fix)
\end{itemize}

\chapter{Conclusion}

speculate on future directions (e.g. concurrency, VIP)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% References:
%%
% If you include some work not referenced in the main text (e.g. using \nocite{}), consider changing "References" to "Bibliography".
%

% \renewcommand to change default "Bibliography" to "References"
\renewcommand{\bibname}{References}
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{References}
\bibliographystyle{plainnat}
\bibliography{thesis}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Appendix:
%%

\appendix

\chapter{Extra Information}
Some more text ...



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Index:
%%
\printthesisindex

\end{document}
